---
title: "Designing a disk cache on mdbx"
description: >
  A draft design for an R equivalent of Python's diskcache, using mdbx instead
  of SQLite — and a working simplified implementation.
output: rmarkdown::html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(mdbx)
```

Python's [diskcache](https://github.com/grantjenks/python-diskcache) is a disk-backed
cache built on SQLite, and [polars-diskcache](https://github.com/lmmx/polars-diskcache)
layers on top of it: the DataFrame goes to a Parquet file, and diskcache tracks
where it went. This article sketches the same thing on mdbx and implements
enough of it to run.

It is a design note, not a package. The point is to show what changes when the
storage engine is an ordered key-value store rather than a relational one.

## What diskcache asks of SQLite

One table and six indexes:

```sql
CREATE TABLE Cache (
  rowid INTEGER PRIMARY KEY, key BLOB, raw INTEGER,
  store_time REAL, expire_time REAL, access_time REAL,
  access_count INTEGER, tag BLOB, size INTEGER,
  mode INTEGER, filename TEXT, value BLOB)

CREATE UNIQUE INDEX Cache_key_raw     ON Cache(key, raw)
CREATE        INDEX Cache_expire_time ON Cache(expire_time)
CREATE        INDEX Cache_store_time  ON Cache(store_time)
CREATE        INDEX Cache_access_time ON Cache(access_time)
CREATE        INDEX Cache_access_count ON Cache(access_count)
CREATE        INDEX Cache_tag_rowid   ON Cache(tag, rowid)
```

Every index exists to answer one question in order and stop early —
`ORDER BY expire_time LIMIT n` for expiry, `ORDER BY access_time LIMIT n` for
least-recently-used eviction, and so on. `filename` is the escape hatch: values
past a size threshold are written as separate files and the row keeps only a
pointer, which is exactly how polars-diskcache stores Parquet.

So SQLite is doing two jobs here: a keyed store, and a set of ordered indexes
that support bounded scans. Neither needs joins, and none of the queries is
relational.

## What that becomes on mdbx

mdbx gives one thing SQLite gives here — **keys in sorted order** — and gives it
directly. A table plus an index becomes a named database whose *key* is the
thing you want to sort by:

| SQLite | mdbx |
| --- | --- |
| `Cache(key) → value` | database `values`, key → payload |
| the metadata columns | database `meta`, key → a serialized record |
| `INDEX (expire_time)` | database `expiry`, key `<expire><key>` |
| `INDEX (access_time)` | database `accessed`, key `<atime><key>` |
| `rowid INTEGER PRIMARY KEY` | `mdbx_dbi_sequence()` |
| `ORDER BY x LIMIT n` | `mdbx_items(db = x, limit = n)` |
| `WHERE x > ?` (resume) | `mdbx_items(db = x, start = k)` |
| `ORDER BY x DESC LIMIT 1` | `mdbx_items(db = x, limit = 1, reverse = TRUE)` |

An index entry is `<8 bytes of time><the key>` rather than just the time, for
two reasons: entries must be unique — two records can expire in the same
millisecond — and having the key in the index means expiry does not need a
second lookup to learn what to delete.

### Byte order is the only order

This is the constraint everything else follows from. mdbx sorts keys as bytes,
so a number has to be encoded such that its byte order *is* its numeric order.
Big-endian, fixed width, does that:

```{r}
be8 <- function(x) writeBin(as.double(x), raw(), size = 8, endian = "big")
un8 <- function(r) readBin(r, "double", size = 8, endian = "big")

times <- c(1, 1000, 1.7e9, 1.7e9 + 1, Inf)
encoded <- vapply(times, function(t) paste(format(be8(t)), collapse = ""), "")
identical(order(encoded), order(times))
```

Note that `Inf` encodes as the largest value of all, which conveniently makes
"never expires" sort after every real deadline — no special case needed.

The usual R instincts do not survive here. `as.character(10) < as.character(9)`,
and little-endian bytes sort by their least significant byte first. Get the
encoding wrong and every ordered scan in this design silently returns the wrong
rows.

## A working draft

```{r}
cache_open <- function(dir, map_size = 256 * 1024^2) {
  dir.create(dir, showWarnings = FALSE, recursive = TRUE)
  dir.create(file.path(dir, "blobs"), showWarnings = FALSE)

  # A cache may lose the last few writes to a crash: it is a cache. SAFE_NOSYNC
  # cannot corrupt the database, only roll it back to the last steady commit.
  env <- mdbx_env_open(file.path(dir, "cache.mdbx"),
                       max_dbs = 8, map_size = map_size, flags = "SAFE_NOSYNC")

  mdbx_with_write(env, function(txn) {
    for (name in c("values", "meta", "expiry", "accessed")) {
      mdbx_dbi_open(txn, name, create = TRUE)
    }
  })

  structure(list(env = env, dir = dir), class = "rcache")
}

# The four databases, resolved against the transaction in hand.
dbs <- function(txn) {
  list(values   = mdbx_dbi_open(txn, "values"),
       meta     = mdbx_dbi_open(txn, "meta"),
       expiry   = mdbx_dbi_open(txn, "expiry"),
       accessed = mdbx_dbi_open(txn, "accessed"))
}

index_key <- function(when, key) c(be8(when), charToRaw(key))
index_of <- function(entry) list(when = un8(entry[1:8]),
                                 key = rawToChar(entry[-(1:8)]))
```

Writing a record touches all four databases, and the transaction makes that one
atomic step — the property SQLite was providing:

```{r}
cache_set <- function(cache, key, value, expire_in = Inf, threshold = 4096) {
  now <- as.numeric(Sys.time())
  expire <- now + expire_in
  payload <- serialize(value, NULL)

  mdbx_with_write(cache$env, function(txn) {
    db <- dbs(txn)
    forget(txn, db, key, cache$dir)  # replacing? drop the old index entries

    # Large values go to a file and the store keeps a pointer -- the same
    # arrangement polars-diskcache uses for Parquet.
    blob <- NULL
    if (length(payload) > threshold) {
      blob <- paste0(format(as.hexmode(sum(utf8ToInt(key) * seq_along(strsplit(key, "")[[1]])))), "-",
                     length(payload), ".rds")
      writeBin(payload, file.path(cache$dir, "blobs", blob))
    } else {
      mdbx_put(txn, key, payload, db = db$values)
    }

    mdbx_put(txn, key, serialize(
      list(stored = now, expire = expire, accessed = now,
           size = length(payload), blob = blob), NULL), db = db$meta)

    mdbx_put(txn, index_key(expire, key), raw(0), db = db$expiry)
    mdbx_put(txn, index_key(now, key), raw(0), db = db$accessed)
  })

  invisible(cache)
}

# Remove a key everywhere it appears. Needs the metadata to find the index
# entries, which is why the index keys carry their timestamp.
forget <- function(txn, db, key, dir) {
  raw_meta <- mdbx_get(txn, key, db = db$meta, as = "raw")
  if (is.null(raw_meta)) return(FALSE)

  meta <- unserialize(raw_meta)
  mdbx_del(txn, index_key(meta$expire, key), db = db$expiry)
  mdbx_del(txn, index_key(meta$accessed, key), db = db$accessed)
  mdbx_del(txn, key, db = db$meta)
  mdbx_del(txn, key, db = db$values)
  if (!is.null(meta$blob)) unlink(file.path(dir, "blobs", meta$blob))
  TRUE
}
```

Reading checks the deadline itself, so an expired entry is invisible before
anything has got round to deleting it:

```{r}
cache_get <- function(cache, key, default = NULL) {
  mdbx_with_read(cache$env, function(txn) {
    db <- dbs(txn)
    raw_meta <- mdbx_get(txn, key, db = db$meta, as = "raw")
    if (is.null(raw_meta)) return(default)

    meta <- unserialize(raw_meta)
    if (meta$expire <= as.numeric(Sys.time())) return(default)

    payload <- if (is.null(meta$blob)) {
      mdbx_get(txn, key, db = db$values, as = "raw")
    } else {
      readBin(file.path(cache$dir, "blobs", meta$blob), raw(), n = meta$size)
    }
    unserialize(payload)
  })
}

cache_delete <- function(cache, key) {
  mdbx_with_write(cache$env, function(txn) forget(txn, dbs(txn), key, cache$dir))
}
```

### Expiry and eviction are the same shape

Both walk an index from its cheap end and stop early. This is the whole reason
the indexes exist, and it is `limit`:

```{r}
# Delete everything already past its deadline, in bounded chunks.
cache_expire <- function(cache, limit = 100) {
  now <- as.numeric(Sys.time())

  mdbx_with_write(cache$env, function(txn) {
    db <- dbs(txn)
    due <- mdbx_items(txn, limit = limit, db = db$expiry, as = "raw")$keys

    removed <- 0
    for (entry in due) {
      at <- index_of(entry)
      if (at$when > now) break        # the rest are in the future
      forget(txn, db, at$key, cache$dir)
      removed <- removed + 1
    }
    removed
  })
}

# Least-recently-used eviction: the oldest access times are the first keys.
cache_evict <- function(cache, n = 10) {
  mdbx_with_write(cache$env, function(txn) {
    db <- dbs(txn)
    oldest <- mdbx_items(txn, limit = n, db = db$accessed, as = "raw")$keys
    for (entry in oldest) forget(txn, db, index_of(entry)$key, cache$dir)
    length(oldest)
  })
}
```

Iteration is resumable, so listing a large cache does not materialise it:

```{r}
cache_keys <- function(cache, chunk = 1000) {
  seen <- character(0)
  from <- NULL

  mdbx_with_read(cache$env, function(txn) {
    db <- dbs(txn)
    repeat {
      keys <- mdbx_keys(txn, limit = chunk, start = from, db = db$meta)
      if (length(keys) == 0) break
      if (!is.null(from)) keys <- keys[-1]   # `start` is inclusive
      seen <<- c(seen, keys)
      if (length(keys) == 0) break
      from <<- keys[length(keys)]
    }
  })
  seen
}

cache_close <- function(cache) {
  mdbx_env_sync(cache$env)
  mdbx_env_close(cache$env)
}
```

## It runs

```{r}
dir <- file.path(tempdir(), "rcache-demo")
cache <- cache_open(dir)

cache_set(cache, "alpha", list(n = 1))
cache_set(cache, "beta", data.frame(x = 1:3))
cache_set(cache, "brief", "gone shortly", expire_in = -1)  # already expired
cache_set(cache, "big", runif(5000))                       # over the threshold

cache_get(cache, "alpha")

cache_get(cache, "beta")

# Expired on read, before anything has deleted it.
cache_get(cache, "brief", default = "(expired)")

# The large one went to a file; the store kept a pointer.
length(cache_get(cache, "big"))
list.files(file.path(dir, "blobs"))
```

```{r}
sort(cache_keys(cache))

# Expiry finds only what is actually due.
cache_expire(cache)

sort(cache_keys(cache))
```

```{r}
# Eviction takes the least recently used first, which is insertion order here.
cache_evict(cache, n = 1)
sort(cache_keys(cache))

cache_close(cache)
```

## What this draft leaves out

It is a sketch, and the gaps are as interesting as the code.

**Access time is never updated on read.** Doing it properly makes every `get()` a
write transaction, which serialises readers against each other — diskcache has
the same problem and solves it with a `statistics` switch and batched updates. A
real implementation would buffer access times and flush them periodically.

**No size accounting or cull-to-limit.** `mdbx_env_info()$file_size` gives the
database size, but the blob directory has to be tracked separately, and
enforcing a byte ceiling means evicting until under it.

**No tags.** A tag index is a fifth database keyed `<tag>\0<key>`, scanned with
`start` for the prefix — the composite-key equivalent of `INDEX (tag, rowid)`.

**Blob names are a toy.** They should be a content hash, and orphaned files need
collecting when a transaction that wrote one rolls back — the store is
transactional, the filesystem is not.

**Blob deletion is not transactional, and this draft gets it wrong on purpose.**
`forget()` calls `unlink()` while the transaction is still open. If a later step
fails, mdbx rolls the metadata back to an entry whose blob file has already
gone, and `cache_get()` on that key then fails to read a file the store still
believes in. That is data loss, not garbage: the reverse of the orphan case
above, and not fixable by collecting unreferenced files. A real implementation
collects the paths to delete during the transaction and unlinks them only after
`mdbx_txn_commit()` returns, accepting orphans on a crash in between — orphans
are recoverable, dangling references are not.

**No concurrent-writer story.** One process writes at a time; a second gets
`MDBX_BUSY` or waits. For a cache that is usually right, but it wants
`flags = "TRY"` and a retry rather than a blocked worker. See
`?"mdbx-concurrency"`.

## What carried over, and what did not

The parts of SQLite this design does *not* need turn out to be most of it: no
schema, no query planner, no SQL, no joins. What it does need — sorted keys,
bounded ordered scans, atomic multi-step writes, and a counter — mdbx provides
directly, and a named database per index is a closer fit to the intent than a
table plus `CREATE INDEX`.

What genuinely gets harder is that **every ordering decision moves into the key
encoding**, where it is your problem rather than the engine's. SQLite knows that
`expire_time` is a number. mdbx knows only bytes, and `be8()` above is the whole
difference between an index that works and one that quietly returns nonsense.
