---
title: "Getting started with mdbx"
description: >
  The patterns you need to use mdbx well: transactions, byte-oriented storage,
  listing, batching, durability, and concurrency.
output: rmarkdown::html_document
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r setup}
library(mdbx)
```

`mdbx` is a thin binding to [libmdbx](https://libmdbx.dqdkfa.ru/), an embedded
transactional key-value store. It is not a database server: there is no process
to start and no connection to make, just a file on disk that one or more
processes map into memory.

This article covers the patterns that matter in practice. For the reference
documentation of any single function, see its help page.

## Environments and transactions

Data lives in an **environment** — a file, opened with `mdbx_env_open()`.

```{r}
path <- file.path(tempdir(), "notes.mdbx")
env <- mdbx_env_open(path)
env
```

Every read and write happens inside a **transaction**. `mdbx_with_write()` opens
one, runs your code, and commits if the block returns — or aborts it if the
block throws, leaving the database exactly as it was.

```{r}
mdbx_with_write(env, function(txn) {
  mdbx_put(txn, "colour", "blue")
  mdbx_put(txn, "size", "large")
})

mdbx_with_read(env, function(txn) mdbx_get(txn, "colour"))
```

That all-or-nothing property is the main thing a transaction buys you. If
something fails part-way, nothing lands:

```{r}
try(mdbx_with_write(env, function(txn) {
  mdbx_put(txn, "colour", "red")
  stop("changed my mind")
}))

# The write above was rolled back with the error.
mdbx_with_read(env, function(txn) mdbx_get(txn, "colour"))
```

When you need to decide at the end whether to keep the writes, drive the
transaction yourself with `mdbx_txn_begin()`:

```{r}
txn <- mdbx_txn_begin(env, write = TRUE)
mdbx_put(txn, "draft", "unfinished")

if (nchar(mdbx_get(txn, "draft")) < 20) {
  mdbx_txn_abort(txn)
} else {
  mdbx_txn_commit(txn)
}

mdbx_txn_state(txn)
```

Aborting is idempotent, so `on.exit(mdbx_txn_abort(txn))` alongside an explicit
commit is a safe belt-and-braces pattern rather than a double-end.

## Keys and values are bytes

libmdbx stores byte strings and records no type. A single string is stored as
its UTF-8 bytes, so `"k"` and `charToRaw("k")` are the same key. Reads decode
back to text by default.

```{r}
mdbx_with_write(env, function(txn) mdbx_put(txn, "ключ", "значение"))
mdbx_with_read(env, function(txn) mdbx_get(txn, "ключ"))
```

Because no type is recorded, decoding is an assumption — but never a silent one.
Anything that is not valid UTF-8 text raises an error naming the fix rather than
returning a mangled string:

```{r}
blob <- serialize(list(retries = 3L), NULL)
mdbx_with_write(env, function(txn) mdbx_put(txn, "config", blob))

try(mdbx_with_read(env, function(txn) mdbx_get(txn, "config")))
```

`as = "raw"` gives you the exact bytes back:

```{r}
mdbx_with_read(env, function(txn) {
  unserialize(mdbx_get(txn, "config", as = "raw"))
})
```

This is why serialization is deliberately left out of the package: `serialize()`
is one choice, but so are JSON, Arrow, and a domain-specific encoding. Pick your
own and store the bytes.

A key that is not there reads as `NULL`, and a stored empty value reads as `""`
— the two never blur.

```{r}
mdbx_with_read(env, function(txn) mdbx_get(txn, "absent"))

mdbx_with_write(env, function(txn) mdbx_put(txn, "blank", ""))
mdbx_with_read(env, function(txn) mdbx_get(txn, "blank"))
```

Pass `default =` when a missing key should read as something else. It is
returned exactly as given, never decoded.

```{r}
mdbx_with_read(env, function(txn) mdbx_get(txn, "absent", default = NA))
```

## Listing what is there

```{r}
mdbx_with_read(env, function(txn) mdbx_keys(txn))
```

Keys come back in **key order** — libmdbx's byte order — not insertion order.
`mdbx_items()` returns keys and values together in a single pass, which is
cheaper than a key list plus a `mdbx_get()` per key.

```{r}
items <- mdbx_with_read(env, function(txn) {
  mdbx_items(txn, limit = 3, as = "raw")
})
names(items)
vapply(items$keys, rawToChar, character(1))
```

Both materialize everything in memory, so an unbounded scan of a database with
more than `mdbx_scan_max` records is refused rather than attempted. Pass an
explicit `limit`, or `limit = Inf` to mean it.

```{r}
mdbx_scan_max
```

## Batch your writes

This is the single most valuable performance habit, and it is worth showing
rather than asserting. The cost of a durable commit is paid *per transaction*,
not per write:

```{r}
words <- sprintf("word-%04d", 1:200)

per_txn <- system.time(
  for (w in words) mdbx_with_write(env, function(txn) mdbx_put(txn, w, "1"))
)[["elapsed"]]

one_txn <- system.time(
  mdbx_with_write(env, function(txn) {
    for (w in words) mdbx_put(txn, w, "2")
  })
)[["elapsed"]]

c(per_transaction = per_txn, single_transaction = one_txn)
```

Each commit flushes to disk, so the gap is really a count of `fsync` calls. How
big it gets depends entirely on the filesystem and hardware — the ratio above is
whatever the machine that built this page managed, and on storage where a
durable flush is expensive it is commonly a hundredfold or more.

Either way the advice is the same, and it costs nothing: group related writes
into one transaction. It is faster *and* it makes them atomic.

The same idea applies to reads, though less dramatically: prefer one
`mdbx_items()` over a loop of `mdbx_get()` calls, because each call crosses the
R/C boundary.

## Durability

Every commit is fully durable by default: a crash at any moment leaves the
database intact. `mdbx_flags()` lists the flags that trade that away.

```{r}
flags <- mdbx_flags()
flags[flags$scope == "env" & flags$settable, c("flag", "runtime")]
```

`"NOMETASYNC"` and `"SAFE_NOSYNC"` can lose recent transactions to a system
crash but cannot corrupt the database. `"UTTERLY_NOSYNC"` can corrupt it beyond
recovery, and is for data you are prepared to regenerate.

Reach for these only after batching, and measure. Relaxing durability speeds up
*commits*; if you have already reduced the number of commits, there is little
left to win. A typical use is a bulk load:

```{r}
bulk <- mdbx_env_open(file.path(tempdir(), "bulk.mdbx"), flags = "SAFE_NOSYNC")

mdbx_with_write(bulk, function(txn) {
  for (i in 1:1000) mdbx_put(txn, sprintf("row-%04d", i), "x")
})

# Nothing was flushed on commit, so make it durable explicitly.
mdbx_env_sync(bulk)
mdbx_env_close(bulk)
```

A single transaction can relax durability for itself alone, leaving the
environment's own setting untouched:

```{r}
txn <- mdbx_txn_begin(env, write = TRUE, flags = "NOSYNC")
mdbx_put(txn, "scratch", "1")
mdbx_txn_commit(txn)
```

## Concurrency

An environment supports **one transaction at a time**: libmdbx binds a
transaction to the thread that began it, and R is single-threaded. Beginning a
second is a clean error, never a hang.

```{r}
txn <- mdbx_txn_begin(env)
try(mdbx_txn_begin(env))
mdbx_txn_abort(txn)
```

Concurrency comes from **separate processes**: many readers and one writer, with
readers never blocking writers. A reader holds the snapshot that existed when it
began, even as another process commits over it.

An environment does not survive a `fork()`, so under `parallel::mclapply()` and
anything else that forks, open the environment *inside* the worker:

```{r}
mdbx_env_close(env)

results <- parallel::mclapply(c("colour", "size"), function(key) {
  worker <- mdbx_env_open(path)
  on.exit(mdbx_env_close(worker))
  mdbx_with_read(worker, function(txn) mdbx_get(txn, key))
}, mc.cores = 1)

unlist(results)
```

Using an inherited one is an error naming the fork rather than a crash. See
`?"mdbx-concurrency"` for the whole contract.

## Inspecting a database

```{r}
env <- mdbx_env_open(path)

stat <- mdbx_env_stat(env)
stat[c("entries", "depth", "pagesize")]

info <- mdbx_env_info(env)
info[c("geo_current", "geo_upper", "numreaders")]
```

`map_size` in `mdbx_env_open()` sets `geo_upper`, the ceiling the file may grow
to. It costs nothing to set generously — the map is virtual address space, not
disk — and a database that hits it raises `MDBX_MAP_FULL`.

`mdbx_version()` reports the bundled library and how it was built.

```{r}
mdbx_version()$describe
```

## Cleaning up

Environments are closed by the garbage collector eventually, but closing
explicitly is what makes it timely. Closing is refused while a transaction is
open, because closing underneath one is undefined behaviour in libmdbx.

```{r}
mdbx_env_close(env)
mdbx_env_is_open(env)
```

With the default single-file layout, a database is the path plus a `-lck`
companion.

```{r}
unlink(c(path, paste0(path, "-lck")))
```
